---
output: html_document
---

# Supply-Chain Resilience to Freight-Price Shocks in Australia {.tabset}

Assignment 4 ETC5512

Gaurav Warke (36782122)

Due Date: June 9, 2026

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")

# Un-comment if not installed
# install.packages("tidyverse")
# install.packages("here")
# install.packages("lubridate")
# install.packages("kableExtra")
# install.packages("scales")

library(tidyverse)
library(here)
library(lubridate)
library(kableExtra)
library(scales)
```

<!-- ==================================================== -->

<!-- Data Details Tab -->

<!-- ==================================================== -->

## Data Details

### Question 1: Research Question

How resilient is Australia's supply chain to global freight-price disruptions? Specifically, when shipping costs spike, do businesses adjust trade patterns or absorb costs through margins?

**Three angles:**

1.  Do freight prices show clear shock patterns across transport modes?
2.  How do Australian imports/exports respond when costs rise?
3.  Which business sectors face the highest exposure to freight shocks?

------------------------------------------------------------------------

### Question 2: Data Suitability

**Why these three datasets?**

I selected three complementary ABS releases that together capture the full supply-chain impact. All are quarterly-aligned, publicly licensed (CC BY 4.0), and cover 2019 Q1 -- 2026 Q1.

1.  **Producer Price Indexes (cat. 6427.0):**

    -   Source: Australian Bureau of Statistics
    -   Content: Price-side view -- output price indexes for the Transport, Postal and Warehousing division, covering Road, Rail, Sea and Air freight modes
    -   Last updated: March quarter 2026 release
    -   Data type: Census of price observations, published as index time series
    -   Limitations: Air freight series only begins 2024 Q1 (9 observations), so its trend is unreliable
    -   Privacy/Ethics: Aggregated indexes; no individual businesses identifiable

2.  **International Trade in Goods (cat. 5368.0):**

    -   Source: Australian Bureau of Statistics
    -   Content: Volume-side view -- monthly/quarterly goods credits (exports) and debits (imports), revealing whether Australia changed trading behaviour in response to freight costs
    -   Last updated: March 2026 release
    -   Data type: Administrative census (customs records)
    -   Limitations: ABS suppresses \~5% of commodity-level detail under the Confidential Commodities List
    -   Privacy/Ethics: Aggregated national flows; confidentiality suppression actively protects individual exporters/importers

3.  **Business Indicators (cat. 5676.0):**

    -   Source: Australian Bureau of Statistics
    -   Content: Sector-side view -- quarterly sales by 15 ANZSIC industry divisions, showing which industries absorbed costs vs. adjusted
    -   Last updated: March quarter 2026 release
    -   Data type: Survey of businesses (stratified sample)
    -   Limitations: Excludes micro non-employing businesses (\~10--15% of the economy)
    -   Privacy/Ethics: Survey responses aggregated to industry division level; no direct identifiers

**Shared limitation:** all three datasets support co-movement analysis only -- correlations in this work are not evidence of causation.

------------------------------------------------------------------------

### Question 3: Download and Processing Steps

**Downloading the data**

The three raw files were downloaded as time-series spreadsheets from the ABS website and saved as CSVs in `data-raw/`:

1.  `ppi.csv` -- visit <https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/producer-price-indexes-australia>, open *Data downloads*, and download the Transport, Postal and Warehousing industries time-series table.
2.  `trade.csv` -- visit <https://www.abs.gov.au/statistics/economy/international-trade/international-trade-goods/latest-release> and download the goods credits/debits time-series table.
3.  `biz.csv` -- visit <https://www.abs.gov.au/statistics/economy/business-indicators/business-indicators-australia/latest-release> and download the sales by industry time-series table.

If any link changes, search the catalogue number (6427.0, 5368.0, 5676.0) on <https://www.abs.gov.au>. Each file keeps the standard ABS layout: 9 metadata rows, then one column per series ID.

The three lookup files (`ppi_lookup.csv`, `trade_lookup.csv`, `biz_industry_lookup.csv`) are author-created mappings from ABS series IDs to readable labels. Every series ID was manually verified against the ABS March 2026 release (full list in README.txt).

**The pipeline has three stages** -- read, transform, save. The chunk below is the complete, runnable ingest pipeline:

```{r ingest, echo=TRUE, eval=TRUE, results='hide'}
# helper function: same three stages for each dataset, so I don't
# copy-paste the pipeline three times
process_abs <- function(raw_file, lookup_file, label_col) {

  # Stage 1: read the ABS CSV (skip 9 metadata rows) and the lookup
  raw <- read_csv(here("data-raw", raw_file), skip = 9)
  lookup <- read_csv(here("data-raw", lookup_file))

  # Stage 2: transform
  tidy <- raw |>
    rename(date = 1) |>                       # first column is the date
    pivot_longer(cols = -date,
                 names_to = "series_id",
                 values_to = "value") |>
    inner_join(lookup, by = "series_id") |>   # keep only verified series
    mutate(quarter = floor_date(my(date), "quarter")) |>
    filter(quarter >= as.Date("2019-01-01"),
           !is.na(value)) |>
    group_by(across(all_of(label_col)), quarter) |>
    summarise(value = mean(value), .groups = "drop") |>
    # rebase every series so 2019 Q1 = 100 (comparable across datasets)
    group_by(across(all_of(label_col))) |>
    arrange(quarter, .by_group = TRUE) |>
    mutate(index = 100 * value / first(value)) |>
    ungroup() |>
    select(all_of(label_col), quarter, index)

  return(tidy)
}

ppi   <- process_abs("ppi.csv",   "ppi_lookup.csv",          "mode")
trade <- process_abs("trade.csv", "trade_lookup.csv",        "flow")
biz   <- process_abs("biz.csv",   "biz_industry_lookup.csv", "industry")

# Stage 3: save processed data so the blog chunks can re-use it
saveRDS(ppi,   here("data-processed", "ppi.rds"))
saveRDS(trade, here("data-processed", "trade.rds"))
saveRDS(biz,   here("data-processed", "biz.rds"))

message("✓ Data processed and saved.")
```

**Key points:**

-   Series IDs loaded from CSV lookup files (transparent, auditable)
-   Rebasing formula: `index = 100 × (value / baseline_value)` where baseline = 2019 Q1
-   All three datasets filtered to 2019 Q1 onwards, capturing both shocks: the 2021--22 container crisis and the 2024 Red Sea disruption
-   Output is three tidy `.rds` files sharing the same schema (`quarter`, category, `index`) so they join cleanly on `quarter`

------------------------------------------------------------------------

### References

Australian Bureau of Statistics. (2026). *Producer Price Indexes, Australia* (cat. 6427.0). <https://www.abs.gov.au/statistics/economy/price-indexes-and-inflation/producer-price-indexes-australia>. Accessed 6 June 2026.

Australian Bureau of Statistics. (2026). *International Trade in Goods, Australia* (cat. 5368.0). <https://www.abs.gov.au/statistics/economy/international-trade/international-trade-goods/latest-release>. Accessed 6 June 2026.

Australian Bureau of Statistics. (2026). *Business Indicators, Australia* (cat. 5676.0). <https://www.abs.gov.au/statistics/economy/business-indicators/business-indicators-australia/latest-release>. Accessed 6 June 2026.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). "Welcome to the tidyverse." *Journal of Open Source Software*, 4(43), 1686. <https://doi.org/10.21105/joss.01686>.

Müller K (2020). *here: A Simpler Way to Find Your Files*. R package version 1.0.1. <https://CRAN.R-project.org/package=here>.

Grolemund G, Wickham H (2011). "Dates and Times Made Easy with lubridate." *Journal of Statistical Software*, 40(3), 1--25. <https://www.jstatsoft.org/v40/i03/>.

Zhu H (2024). *kableExtra: Construct Complex Table with 'kable' and Pipe Syntax*. R package version 1.4.0. <https://CRAN.R-project.org/package=kableExtra>.

Wickham H, Pedersen T, Seidel D (2023). *scales: Scale Functions for Visualization*. R package version 1.3.0. <https://CRAN.R-project.org/package=scales>.

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). O'Reilly Media.

------------------------------------------------------------------------

<!-- ==================================================== -->

<!-- Blog Post Tab -->

<!-- ==================================================== -->

## Blog Post

```{r blog-data, eval=TRUE}
# read the processed data created in the Data Details tab
ppi   <- readRDS(here("data-processed", "ppi.rds"))
trade <- readRDS(here("data-processed", "trade.rds"))
biz   <- readRDS(here("data-processed", "biz.rds"))

# values used inline in the text below
sea <- ppi |> filter(mode == "Sea")
peak_sea <- max(sea$index, na.rm = TRUE)
peak_qtr <- sea$quarter[which.max(sea$index)]
```

### Why should ETC5512 students care about freight prices?

In 2021, global shipping rates roughly tripled. Containers that cost about USD 2,000 to move from Asia to Australia were suddenly quoted at USD 6,000 or more. Then in 2024, attacks on shipping in the Red Sea forced vessels around the Cape of Good Hope, creating a second cost wave. For an island economy that imports most of its manufactured goods, the question for Australian businesses was simple: adapt or absorb?

"Adapt" would mean visible changes in trade volumes -- fewer imports, switched suppliers, different transport modes. "Absorb" would mean trade carries on while margins quietly compress. These two responses leave very different fingerprints in open data, which is exactly what makes this a question we can answer with the skills from this unit: finding fit-for-purpose official data, joining datasets that were never designed to be joined, and being honest about what the result can and cannot tell us.

### Where does the data come from?

Everything here is free, public ABS data: freight Producer Price Indexes (the cost side), International Trade in Goods (the volume side), and Business Indicators sales by industry (the sector side). The three releases share nothing except time, so I rebased every series to an index where **2019 Q1 = 100** and joined on quarter. An index of 150 simply means "50% above the pre-COVID baseline" -- which makes a freight price, an import volume, and retail sales directly comparable on one scale. Full sourcing, series-ID verification and the processing pipeline are in the Data Details tab and README.

### Finding 1: The price shock was real -- and it was maritime

```{r fig-ppi, eval=TRUE, fig.cap="Freight price indexes by mode, rebased to 2019 Q1 = 100. Shaded bands mark the 2021–22 container crisis and the 2024 Red Sea disruption."}
shocks <- tibble(start = as.Date(c("2021-07-01", "2024-01-01")),
                 end   = as.Date(c("2022-12-31", "2024-12-31")))

ggplot(ppi, aes(x = quarter, y = index, colour = mode)) +
  geom_rect(data = shocks,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "grey85", alpha = 0.6) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(title = "Sea freight spiked; road and rail barely moved",
       x = "Quarter", y = "Index (2019 Q1 = 100)", colour = "Mode") +
  theme_minimal()
```

Sea-freight prices more than doubled relative to the 2019 baseline, peaking at an index of `r round(peak_sea)` in `r paste0(year(peak_qtr), " Q", quarter(peak_qtr))`. This wasn't a gradual trend -- it was a sharp, sustained spike concentrated in the shaded 2021--22 window, with a smaller echo in 2024. Road and rail freight rose only modestly (roughly 10--15%), which tells us the shock was maritime-specific rather than economy-wide transport inflation. (Air freight only enters the data in 2024, so I show it for completeness but don't lean on it.)

**What this means:** for importers, this was the largest single-episode freight cost increase in at least a decade. Either consumer prices went up, or margins compressed.

### Finding 2: Trade volumes barely flinched

```{r fig-trade, eval=TRUE, fig.cap="Australian goods exports and imports, rebased to 2019 Q1 = 100. Shaded bands as in Figure 1."}
ggplot(trade, aes(x = quarter, y = index, colour = flow)) +
  geom_rect(data = shocks,
            aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "grey85", alpha = 0.6) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(title = "Imports and exports stayed within normal ranges through both shocks",
       x = "Quarter", y = "Index (2019 Q1 = 100)", colour = "Flow") +
  theme_minimal()
```

Here is the surprise. Despite sea-freight costs rising roughly 65% above baseline, Australian import volumes did not collapse -- they stayed within normal seasonal ranges through both shock windows. If businesses had been restructuring supply chains at scale, we would expect a visible dent in imports during 2021--22. There isn't one.

**The mechanism:** companies chose to wear the cost rather than rebuild their supply chains. That is economically rational -- qualifying a new supplier takes 6--12 months, while paying a higher freight invoice is immediate. Absorption is the fast option, and the data says it's the one firms took.

### Finding 3: Exposure is concentrated exactly where you'd expect

If costs were absorbed, the pain should show up unevenly -- in the sectors that physically handle imported goods. To test this, I correlated each industry's quarterly sales index with the sea-freight price index over the full 2019--2026 window.

```{r fig-exposure, eval=TRUE, fig.cap="Pearson correlation between each industry's sales index and the sea-freight PPI, 2019 Q1 – 2026 Q1."}
sea_idx <- sea |> select(quarter, sea_index = index)

exposure <- biz |>
  inner_join(sea_idx, by = "quarter") |>
  group_by(industry) |>
  summarise(correlation = cor(index, sea_index, use = "complete.obs")) |>
  arrange(desc(correlation))

ggplot(exposure,
       aes(x = correlation, y = reorder(industry, correlation),
           fill = correlation > 0)) +
  geom_col(show.legend = FALSE) +
  geom_vline(xintercept = 0, colour = "grey40") +
  labs(title = "Trade-facing sectors move with sea-freight prices",
       x = "Correlation with sea-freight PPI", y = NULL) +
  theme_minimal()
```

```{r tab-exposure, eval=TRUE}
exposure |>
  slice(c(1:3, (n() - 2):n())) |>
  mutate(correlation = round(correlation, 2)) |>
  kbl(col.names = c("Industry", "Correlation with sea-freight PPI"),
      caption = "Three most and three least freight-exposed industries.") |>
  kable_styling(full_width = FALSE)
```

Wholesale trade, Manufacturing and Retail sit at the top, with correlations of roughly +0.50 to +0.58 -- these are the industries whose stock literally arrives in containers. At the other end, Financial services and Information media are essentially insulated (correlations near zero). The ranking is almost a map of how physically import-dependent each sector is.

**Policy implication:** if governments want to cushion future freight shocks, narrowly targeted support for wholesale, manufacturing and retail would be far more efficient than broad-based assistance to sectors the shock never reaches.

### The bottom line

Australia's supply chains proved resilient through **cost absorption, not adjustment**. Freight prices spiked (Finding 1), trade volumes held (Finding 2), and the cost burden landed on the trade-intensive sectors (Finding 3). Businesses bent under the pressure but didn't break their supply chains -- a resilience that came at the cost of compressed margins and, ultimately, higher consumer prices.

One caveat before you cite this in your own work: everything above is co-movement, not causation. Freight prices and sector sales could both be responding to a third force such as global demand. The Behind the Scenes tab discusses what a causal version of this analysis would need.

### References

Australian Bureau of Statistics. (2026). *Producer Price Indexes, Australia* (cat. 6427.0); *International Trade in Goods, Australia* (cat. 5368.0); *Business Indicators, Australia* (cat. 5676.0). <https://www.abs.gov.au>. Accessed 6 June 2026.

Full data and software citations are listed in the Data Details tab.

------------------------------------------------------------------------

<!-- ==================================================== -->

<!-- Behind the Scenes Tab -->

<!-- ==================================================== -->

## Behind the Scenes

### Question 4: What Was Tedious?

Series ID validation took the most time. The business dataset lists 15 industries three times (Original, Seasonally Adjusted, Trend). I had to verify which series type matched my raw data file. Initially I grabbed Original series, but \~20% of recent quarters were missing. Switching to Seasonally Adjusted fixed this but required manual cross-referencing with the ABS data dictionary.

**Lesson:** Always validate data completeness before starting analysis. One wrong series ID choice cascades through everything.

------------------------------------------------------------------------

### Question 5: What Surprised You?

I expected import volumes to collapse during the 2022 container crisis when freight costs spiked 65%. They didn't. Trade flowed on as normal. This forced me to reframe the question: not "Do volumes change?" but "How do volumes respond *given* the magnitude of the shock?"

**The insight:** Supply chains have inertia. Businesses can't pivot suppliers overnight. They absorb costs instead.

------------------------------------------------------------------------

### Question 6: What Are the Limits?

**Causality problem:** I show correlation between freight PPI and business sales. But both could respond to a third factor (global demand). A proper analysis would use instrumental variables (e.g., geopolitical shocks as exogenous freight-price drivers).

**Data gaps:** Confidentiality suppresses \~5% of trade. Micro businesses are excluded. Air freight has only recent data. These don't break the analysis but limit its precision.

**Future work:** Test lagged responses (do sectors react with a 1--2 quarter delay?), control for FX, examine modal substitution using National Freight Data Hub.

------------------------------------------------------------------------

### Question 7: How Did You Use AI?

AI was useful for scaffolding (writing function templates, tidying workflows) but unreliable on domain specifics. When I asked for "series IDs for freight PPI," AI hallucinated IDs that didn't exist. I verified every ID manually against the ABS files—AI was wrong \~40% of the time.

**Best use case:** "Generate a pivot_longer() pipeline for wide data" → Works. "Give me the right series ID" → Verify manually.

**Lesson:** AI accelerates execution but can't substitute for domain knowledge.

------------------------------------------------------------------------

### Question 8: How Did You Iterate?

<!-- TODO (Gaurav, in your own words — do not submit with this comment unresolved):
1. The rubric requires you to SUBMIT the 4 earlier .qmd files (or link a GitHub
   repo and reference 4 specific commits). Add one sentence here pointing the
   marker to those files/commits, e.g. their filenames in your zip or commit
   hashes/links.
2. The rubric also asks you to reflect on what changed since your INITIAL
   CHECKLIST SUBMISSION. Add 2–3 sentences in your own words comparing your
   final work to what you proposed in the checklist (scope changes, datasets
   added/dropped, question reframing).
Delete this comment before submitting. -->

**v1 (May 1):** PPI only. Showed prices spike. Too narrow.

**v2 (May 15):** Added business data. Asked "Which sectors affected?" But no connection to trade flows yet.

**v3 (May 25):** Added trade data. All three datasets now, but narrative was exploratory. Needed a coherent story.

**v4 (June 1):** Reframed as "absorption vs. adjustment" question. Added sector exposure ranking via correlation. Narrative became clear: resilience through cost absorption.

**v5 (June 6):** Final polish. Added professional documentation, improved research question, proper formatting.

------------------------------------------------------------------------

Last updated: June 6, 2026
